---
title: "Donor Retention Project"
author: "Artea Myftiu"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,   # hide code by default (can be unfolded)
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4
)
set.seed(2025)
```

```{r libraries}
library(tidyverse)
library(lubridate)
library(tidymodels)
library(pROC)
library(rpart)
library(ranger)
library(xgboost)
library(yardstick)
library(baguette)
library(gt)
```

# 1. Executive Summary

This project builds predictive models to estimate whether each donor will
donate again in 2018. Using historical donation behavior from multiple
years, we:

- Construct features summarizing **recency, frequency and monetary** behavior.
- Train several models:
  - Logistic regression (full and forward stepwise–selected)
  - Decision trees (default and tuned)
  - Bagged trees
  - Random forest
  - Gradient boosting (XGBoost)
- Evaluate models on a hold-out **2018 test set** using:
  - AUC
  - Sensitivity
  - Specificity

We then translate the best predictive model into **business impact** by
simulating different mailing strategies (what % of donors to contact) and
computing cost, revenue, and net profit. We also build a **lift chart**
to show how much better the model is than a random strategy.

**Key high-level findings:**

- The **stepwise logistic regression** achieves the highest test AUC
  (≈ 0.693), slightly better than the random forest and other tree-based
  ensembles.
- The business problem is **imbalanced but not extreme**: only about
  20–21% of donors donate in the prediction year.
- Targeting only the **top-scoring donors** (e.g. top 60–80%) generates
  almost the same revenue as mailing everyone, with lower cost and
  slightly higher net profit.
- The model helps the charity **spend mailing budget more efficiently**
  and suggests a move away from the naive “mail everyone” strategy.

# 2. Business Problem & Data

The organization wants to **predict which donors are most likely to give
again** so that marketing mailers can be focused on high-probability
donors. Mailing everyone is expensive and wastes budget on low-probability
donors.

We use five data files:

- `donors.csv` — static donor demographics and segmentation
- `gifts.csv` — donation history (donor, date, amount)
- `basetable_2017.csv` — engineered features and target for 2017
- `basetable_2018.csv` — engineered features and target for 2018 (test)
- `feature_catalog.csv` — documentation of the engineered features

## 2.1 Load Data

```{r load-data}
donors          <- read_csv("donors.csv")
gifts           <- read_csv("gifts.csv")
basetable_2017  <- read_csv("basetable_2017.csv")
basetable_2018  <- read_csv("basetable_2018.csv")
feature_catalog <- read_csv("feature_catalog.csv")
```

## 2.2 Exploratory Data Analysis

We first explore the donor universe and the basetables to understand
the size of the dataset, missing values, and donation behavior.

```{r eda}
# Donor demographics
glimpse(donors)
summary(donors)

# Clean gifts and inspect
gifts <- gifts %>%
  mutate(
    donor_id = as.integer(id),
    date     = as.Date(date)
  )

glimpse(gifts)
summary(gifts)

# Date range of donations and number of unique donors
date_range <- range(gifts$date, na.rm = TRUE)
unique_donors <- length(unique(gifts$donor_id))

date_range
unique_donors

# Basetables
glimpse(basetable_2017)
glimpse(basetable_2018)

summary(basetable_2017$target_donated)
summary(basetable_2018$target_donated)

na_2017 <- colSums(is.na(basetable_2017))
na_2018 <- colSums(is.na(basetable_2018))

na_2017
na_2018

head(feature_catalog)
```

### 2.3 Question 1: Data Understanding

**a) What is the date range of the donation history? How many years of data are available?**  
From the gifts data, the donation dates range from **2012-01-01** to
**2018-12-31**, which is roughly **7 years** of donation history.

**b) What percentage of donors have missing demographic information (age, gender, location)?**  
In our basetables, the key demographic fields (`age`, `gender`,
`country`, `region`, `segment`) have **0 missing values**, so the
percentage of donors with missing demographic information is effectively
**0%**.

**c) Overall donation response rate across all historical campaigns**  
In `basetable_2017`, the mean of `target_donated` ≈ **0.212**  
In `basetable_2018`, the mean of `target_donated` ≈ **0.204**  

So the overall response rate is around **20–21%**, meaning **most donors
do not donate in the prediction window**. This confirms that the
business problem is a **moderately imbalanced classification problem**
where we care about lifting the probability of donation above this
baseline.

**d) Data quality issues before modeling**  

- **Missing values:** `avg_days_between` has some NAs (12 in 2017, 6 in
  2018) that need to be imputed.
- **Imbalanced target:** Only ~20% positive cases; evaluation metrics
  should emphasize AUC and lift rather than accuracy alone.
- **Monetary outliers:** Donation amounts vary widely; a few very large
  gifts may influence models if not handled carefully.
- **Zero recent activity but non-zero lifetime:** Many donors have no
  donations in the recent 12-month window but older lifetime donations –
  this is meaningful but should be checked for consistency.

# 3. Data Preparation

We clean a few variables and make sure categorical variables are treated
as factors.

```{r prep}
# Fix avg_days_between NAs by median imputation
for (df_name in c("basetable_2017", "basetable_2018")) {
  df <- get(df_name)
  if ("avg_days_between" %in% names(df)) {
    med_val <- median(df$avg_days_between, na.rm = TRUE)
    df$avg_days_between[is.na(df$avg_days_between)] <- med_val
  }
  assign(df_name, df)
}

# Ensure categorical variables are factors (if they exist)
cat_vars <- intersect(
  c("gender", "country", "region", "segment"),
  names(basetable_2017)
)

basetable_2017 <- basetable_2017 %>%
  mutate(across(all_of(cat_vars), as.factor))

basetable_2018 <- basetable_2018 %>%
  mutate(across(all_of(cat_vars), as.factor))
```

### 3.1 Conceptual Questions on Windows & Population

> These questions relate to the general course framework (observation
> dates, lookback/prediction windows, population selection), not only
> this specific dataset.

#### 3.1.1 Observation Dates & Windows

**a) Why choose quarterly observation dates vs monthly or annual?**  
Quarterly observation dates are a **compromise**:

- Compared to **annual**, they give **more training examples** and
  capture shorter-term behavior changes.
- Compared to **monthly**, they avoid producing too many highly
  correlated samples and reduce computation.  
Overall, quarterly points capture dynamics without exploding the number
of rows.

**b) Why must the 365-day lookback not overlap with the 90-day prediction window?**  
If the windows overlap, features would use information from the **future
relative to the prediction** – this is a **timeline violation / data
leakage**. The model would then look artificially good in evaluation but
perform poorly in production.

**c) Timeline diagram (conceptual)**  

For an observation date of 2023-07-01:

- Lookback window: **2022-07-02 to 2023-07-01** (365 days)
- Observation date: **2023-07-01** (vertical line)
- Prediction window: **2023-07-02 to 2023-09-30** (next 90 days)

All features are computed using data **strictly before** the observation
date; the target is whether a donation occurs **strictly after** it.

**d) Number of observation points and training examples**  
If we pick, for example, **4 quarterly observation dates per year** over
**2 years**, that gives **8 observation points**.  
If each yields **10,000 eligible donors**, we get about  
**8 × 10,000 = 80,000** training examples.

#### 3.1.2 Population Selection

**a) Why require at least one donation in the lookback window?**  
We want to model **retention** of active donors. Requiring at least one
donation ensures we exclude people with **no recent activity** (cold
leads). This focuses the model on donors where mailing decisions are
most relevant.

**b) 8,000 eligible donors vs 50,000 total**  
The missing 42,000 donors are:

- New donors with **no history in the lookback window**
- Dormant donors whose last donation is too old
- Donors filtered out by business rules  

This is expected and not necessarily bad, as long as the definition of
“eligible” matches the business goal.

**c) Timeline compliance**  
“Timeline compliant” means **no feature uses information from the
future** relative to the observation date.

Example violation: using a **lifetime donation total computed as of
2024-01-01** to predict behavior for an observation date in 2023. This
includes data from **after** the observation point and would leak
future information into the model.

#### 3.1.3 Target Definition

**a) 9.5% response rate – balanced or imbalanced?**  
A response rate of **9.5%** is **clearly imbalanced** (far more
non-donors than donors). This implies we should:

- Focus on **AUC, PR curves, and lift**, rather than accuracy.
- Consider models that handle imbalance better (e.g., tree ensembles).
- Be careful with decision thresholds (0.5 may not be optimal).

**b) Why must prediction window start after the observation date?**  
If the prediction window includes the observation date, we risk counting
donations that were already known at the time of prediction, again
causing **data leakage**. The target must reflect **future behavior**,
not current or past events.

#### 3.1.4 Feature Engineering (RFM etc.)

**a) Intuition behind RFM components**

- **Recency:** Donors who gave **recently** are much more likely to give
  again soon; long gaps usually mean engagement is fading.
- **Frequency:** Donors who give **often** show strong engagement and are
  more likely to respond again.
- **Monetary:** Donors who give **larger amounts** often care more or
  have higher capacity to give; they tend to be higher-value targets.

**b) Why multiple time windows (30, 90, 180, 365 days)?**  

Different windows capture **different patterns**:

- 30 days: very **short-term spikes** or campaign responses
- 90 days: **recent engagement** (quarterly behavior)
- 180 days: medium-term consistency
- 365 days: **overall annual activity**

Some donors give only once per year; others give frequently in bursts.
Multiple windows let the model distinguish these.

**c) Trend indicator formula and interpretation**  

A simple trend measure is:

Trend = (donations last 90 days − donations 91–365 days ago) /
        (donations 91–365 days ago + ε)



- **Positive** trend: activity is **increasing** recently.
- **Negative** trend: donor was more active in the **past** than
  recently (possible churn).

**d) Three additional useful features**

- **Channel mix** (e.g., share of online vs offline gifts) – donors who
  use convenient channels may respond better to certain campaigns.
- **Time since first donation** – distinguishes **very new** donors from
  long-time loyal donors.
- **Number of distinct years with a donation** – measures **loyalty over
  time**, not just recent frequency.

#### 3.1.5 Basetable Quality Control

**a) Code to check basic basetable quality**

Conceptually, we want to ensure no feature uses dates after the
observation date or inside the prediction window. In this dataset,
features are already pre-computed, so we mainly check:

```{r basetable-qc}
# Example QC summary

nrows_2017 <- nrow(basetable_2017)
nrows_2018 <- nrow(basetable_2018)

unique_donors_2017 <- n_distinct(basetable_2017$donor_id)
unique_donors_2018 <- n_distinct(basetable_2018$donor_id)

nrows_2017
nrows_2018
unique_donors_2017
unique_donors_2018

# Target distribution
prop.table(table(basetable_2017$target_donated))
prop.table(table(basetable_2018$target_donated))

# Missing value percentages
missing_pct_2017 <- colSums(is.na(basetable_2017)) / nrow(basetable_2017)
missing_pct_2018 <- colSums(is.na(basetable_2018)) / nrow(basetable_2018)

missing_pct_2017
missing_pct_2018
```

**b) Descriptive statistics for basetable**

- `basetable_2017`:
  - Rows: **900** donor-observation pairs
  - Unique donors: **900**
  - Observation dates: pre-aggregated (one row per donor)
  - Target positive rate (`target_donated`): ≈ **21.2%**

- `basetable_2018`:
  - Rows: **651** donor-observation pairs
  - Unique donors: **651**
  - Target positive rate: ≈ **20.4%**

Missing values:

- Most features have **0% missing**.
- `avg_days_between` had some NAs, which we imputed with the median.

**c) 15% missing email_open_rate – why and what to do?**

- **Why:** Some donors may never have been emailed, or email tracking
  wasn’t available earlier → missing by design / not measured.
- **Two strategies:**
  1. Create a **“missing” category / flag** to let the model learn if
     “no email data” itself is informative.
  2. Impute using **group-based averages** (e.g., by segment, country)
     and track a separate indicator that the value was imputed.

# 4. Logistic Regression Models

We first build logistic regression models on the 2017 basetable. The
goal is to predict `target_donated` (donated again next year).

## 4.1 Full Logistic Model

```{r logit-full}
# Split 2017 into train/validation
set.seed(2025)
n_2017    <- nrow(basetable_2017)
train_idx <- sample(seq_len(n_2017), size = floor(0.75 * n_2017))

train_2017 <- basetable_2017[train_idx, ]
valid_2017 <- basetable_2017[-train_idx, ]

# Full formula: all predictors except donor_id & target_amount
full_formula <- as.formula("target_donated ~ . - donor_id - target_amount")

logit_full <- glm(
  full_formula,
  data   = train_2017,
  family = binomial
)

summary(logit_full)

valid_probs_full <- predict(logit_full, newdata = valid_2017, type = "response")
roc_full         <- roc(valid_2017$target_donated, valid_probs_full)
auc_full         <- auc(roc_full)
auc_full
```

## 4.2 Forward Stepwise Selection (AUC-based)

```{r logit-stepwise}
forward_stepwise_auc <- function(
    candidate_vars,
    data_train,
    data_valid,
    target  = "target_donated",
    max_vars = 10
) {
  selected_vars  <- character(0)
  remaining_vars <- candidate_vars
  best_auc       <- 0.5
  history <- tibble(
    step           = integer(),
    variable_added = character(),
    auc            = numeric()
  )

  for (step in seq_len(max_vars)) {
    step_results <- map_dfr(remaining_vars, function(v) {
      vars_try <- c(selected_vars, v)
      f <- as.formula(
        paste(target, "~", paste(vars_try, collapse = " + "))
      )
      fit   <- glm(f, data = data_train, family = binomial)
      preds <- predict(fit, newdata = data_valid, type = "response")
      r <- tryCatch(roc(data_valid[[target]], preds), error = function(e) NA)
      auc_val <- if (any(is.na(r))) NA_real_ else as.numeric(auc(r))
      tibble(variable = v, auc = auc_val)
    })

    step_results <- step_results %>%
      filter(!is.na(auc))

    if (nrow(step_results) == 0) break

    best_idx  <- which.max(step_results$auc)
    best_step <- step_results[best_idx, , drop = FALSE]

    if (best_step$auc <= best_auc + 1e-4) {
      break
    }

    best_auc <- best_step$auc
    best_var <- best_step$variable
    selected_vars  <- c(selected_vars, best_var)
    remaining_vars <- setdiff(remaining_vars, best_var)

    history <- history %>%
      add_row(
        step           = step,
        variable_added = best_var,
        auc            = best_auc
      )
  }

  list(
    selected_vars       = selected_vars,
    performance_history = history
  )
}

candidate_vars <- setdiff(
  names(train_2017),
  c("donor_id", "target_amount", "target_donated")
)

fs_result <- forward_stepwise_auc(
  candidate_vars = candidate_vars,
  data_train     = train_2017,
  data_valid     = valid_2017,
  target         = "target_donated",
  max_vars       = 10
)

fs_result$selected_vars
fs_result$performance_history

# Fit selected logistic model
if (length(fs_result$selected_vars) > 0) {
  selected_formula <- as.formula(
    paste("target_donated ~", paste(fs_result$selected_vars, collapse = " + "))
  )
  logit_selected <- glm(
    selected_formula,
    data   = train_2017,
    family = binomial
  )

  valid_probs_sel <- predict(logit_selected, newdata = valid_2017, type = "response")
  roc_sel         <- roc(valid_2017$target_donated, valid_probs_sel)
  auc_sel         <- auc(roc_sel)
  auc_sel
} else {
  logit_selected <- NULL
  auc_sel        <- NA_real_
}
```

### 4.3 Logistic Model Interpretation & Variable Selection


- Full logistic validation AUC ≈ **0.564**
- Stepwise logistic validation AUC ≈ **0.615**
- On 2018 test set:
  - Full logistic AUC ≈ **0.669**
  - Selected logistic AUC ≈ **0.693**

#### Question 3.1: Initial Model Interpretation

**a)** AUC of the full logistic model on the test set is ≈ **0.669**.  
According to the guideline (0.7 acceptable, 0.8 good), this is just
below the “acceptable” threshold, so performance is **moderate but not
strong**.

**b)** The variables with the largest **positive** coefficients (in
magnitude) include things like:

- Being in higher value **segments** (e.g., Gold / Silver)
- Higher **recent monetary totals**
- Higher **recent frequencies**

These indicate donors who give more often and/or larger amounts are more
likely to give again.

The variables with the largest **negative** coefficients include:

- Very long **recency** (larger days since last donation)
- Some combinations of low frequency and low monetary amounts
- Features indicating low or declining engagement  

In short: **recent, frequent and higher monetary donors** are predicted
to respond again; donors with long inactivity are less likely.

**c)** McFadden’s pseudo \(R^2\) is usually relatively low for logistic
models in marketing; a value around, say, 0.1–0.2 means the model
explains **a modest but meaningful share** of the variation in donation
behavior. It confirms that **donor behavior is noisy**, and even good
models will not be perfectly predictive.

#### Question 3.2: Variable Selection (Forward Stepwise)

**a)** The first selected variable was **`monetary_max_12m`**.  
This makes sense: donors who had at least one **large gift in the last
year** are strongly indicative of both capacity and engagement.

**b)** As we add variables, the AUC improves at first (e.g., 0.586,
0.602, 0.609, 0.613, 0.615). After a few steps, the AUC gain from
adding more variables becomes **very small** (diminishing returns after
~5 variables).

**c)** The stepwise model uses **far fewer variables** than the full
model (e.g., 5–7 vs ~25) but achieves **better AUC** on the test set
(0.693 vs 0.669). This illustrates the **bias–variance tradeoff**: a
simpler model with the strongest predictors can generalize better and
avoid overfitting noise in the full model.

# 5. Tree-Based Models (tidymodels)

We now move to tree-based models using tidymodels: a single decision
tree, tuned tree, bagging, random forest, and gradient boosting.

## 5.1 Prepare Data and Resampling

```{r trees-prep}
data_2017 <- basetable_2017 %>%
  mutate(
    target = factor(if_else(target_donated == 1, "yes", "no"))
  )

data_2018 <- basetable_2018 %>%
  mutate(
    target = factor(if_else(target_donated == 1, "yes", "no"))
  )

set.seed(2025)
data_split <- initial_split(data_2017, prop = 0.75, strata = target)
train_tree <- training(data_split)
valid_tree <- testing(data_split)
test_tree  <- data_2018

tree_rec <- recipe(target ~ ., data = train_tree) %>%
  update_role(donor_id, target_donated, target_amount, new_role = "ID") %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

set.seed(2025)
cv_folds <- vfold_cv(train_tree, v = 5, strata = target)
```

## 5.2 Single Decision Tree

```{r tree-single}
tree_spec <- decision_tree(
  mode           = "classification",
  cost_complexity = 0.01,
  tree_depth      = 10,
  min_n           = 20
) %>%
  set_engine("rpart")

tree_wf <- workflow() %>%
  add_model(tree_spec) %>%
  add_recipe(tree_rec)

tree_fit <- tree_wf %>%
  fit(data = train_tree)

tree_pred_test <- predict(tree_fit, new_data = test_tree, type = "prob")

tree_auc_test <- auc(
  roc(
    as.numeric(test_tree$target == "yes"),
    tree_pred_test$.pred_yes
  )
)
tree_auc_test
```

## 5.3 Tuned Decision Tree

```{r tree-tuned}
tree_tune_spec <- decision_tree(
  mode           = "classification",
  cost_complexity = tune(),
  tree_depth      = tune(),
  min_n           = tune()
) %>%
  set_engine("rpart")

tree_grid <- grid_regular(
  cost_complexity(),
  tree_depth(range = c(3L, 15L)),
  min_n(range = c(5L, 50L)),
  levels = 4
)

tree_tune_wf <- workflow() %>%
  add_model(tree_tune_spec) %>%
  add_recipe(tree_rec)

set.seed(2025)
tree_tune_res <- tune_grid(
  tree_tune_wf,
  resamples = cv_folds,
  grid      = tree_grid,
  metrics   = metric_set(roc_auc)
)

best_tree_params <- select_best(tree_tune_res, metric = "roc_auc")
best_tree_params

best_tree_wf <- finalize_workflow(tree_tune_wf, best_tree_params) %>%
  fit(data = train_tree)

best_tree_pred_test <- predict(best_tree_wf, new_data = test_tree, type = "prob")

best_tree_auc_test <- auc(
  roc(
    as.numeric(test_tree$target == "yes"),
    best_tree_pred_test$.pred_yes
  )
)
best_tree_auc_test
```

## 5.4 Bagging (Bagged Trees)

```{r bagging}
bag_spec <- bag_tree(
  mode           = "classification",
  cost_complexity = 0.01,
  tree_depth      = 15,
  min_n           = 10
) %>%
  set_engine("rpart", times = 50)

bag_wf <- workflow() %>%
  add_model(bag_spec) %>%
  add_recipe(tree_rec)

bag_fit <- bag_wf %>%
  fit(data = train_tree)

bag_pred_test <- predict(bag_fit, new_data = test_tree, type = "prob")

bag_auc_test <- auc(
  roc(
    as.numeric(test_tree$target == "yes"),
    bag_pred_test$.pred_yes
  )
)
bag_auc_test
```

## 5.5 Random Forest

```{r rf}
rf_tune_spec <- rand_forest(
  mode  = "classification",
  mtry  = tune(),
  trees = 500,
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity")

# Rough upper bound for mtry
p_train <- ncol(model.matrix(target ~ ., data = train_tree)) - 1

rf_grid <- grid_regular(
  mtry(range = c(5L, min(20L, p_train))),
  min_n(range = c(5L, 50L)),
  levels = 4
)

rf_wf <- workflow() %>%
  add_model(rf_tune_spec) %>%
  add_recipe(tree_rec)

set.seed(2025)
rf_tune_res <- tune_grid(
  rf_wf,
  resamples = cv_folds,
  grid      = rf_grid,
  metrics   = metric_set(roc_auc)
)

best_rf_params <- select_best(rf_tune_res, metric = "roc_auc")
best_rf_params

best_rf_wf <- finalize_workflow(rf_wf, best_rf_params) %>%
  fit(data = train_tree)

rf_pred_test <- predict(best_rf_wf, new_data = test_tree, type = "prob")

rf_auc_test <- auc(
  roc(
    as.numeric(test_tree$target == "yes"),
    rf_pred_test$.pred_yes
  )
)
rf_auc_test

# Variable importance
rf_fit_obj <- extract_fit_parsnip(best_rf_wf)$fit
rf_varimp <- as.data.frame(rf_fit_obj$variable.importance) %>%
  rownames_to_column("variable") %>%
  rename(importance = 2) %>%
  arrange(desc(importance))

rf_varimp_top15 <- head(rf_varimp, 15)

rf_varimp_top15

rf_varimp_top15 %>%
  ggplot(aes(x = reorder(variable, importance), y = importance)) +
  geom_col() +
  coord_flip() +
  ggtitle("Top 15 Variable Importance - Random Forest") +
  xlab("Variable") + ylab("Importance")
```

## 5.6 Gradient Boosting (XGBoost)

```{r xgb}
boost_tune_spec <- boost_tree(
  mode       = "classification",
  trees      = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  min_n      = tune()
) %>%
  set_engine("xgboost")

boost_grid <- grid_latin_hypercube(
  trees(range = c(100L, 700L)),
  tree_depth(range = c(3L, 8L)),
  learn_rate(range = c(0.01, 0.3)),
  min_n(range = c(5L, 50L)),
  size = 20
)

boost_wf <- workflow() %>%
  add_model(boost_tune_spec) %>%
  add_recipe(tree_rec)

set.seed(2025)
boost_tune_res <- tune_grid(
  boost_wf,
  resamples = cv_folds,
  grid      = boost_grid,
  metrics   = metric_set(roc_auc)
)

best_boost_params <- select_best(boost_tune_res, metric = "roc_auc")
best_boost_params

best_boost_wf <- finalize_workflow(boost_wf, best_boost_params) %>%
  fit(data = train_tree)

boost_pred_test <- predict(best_boost_wf, new_data = test_tree, type = "prob")

boost_auc_test <- auc(
  roc(
    as.numeric(test_tree$target == "yes"),
    boost_pred_test$.pred_yes
  )
)
boost_auc_test
```

### 5.7 Tree & Ensemble Questions


- Single decision tree AUC (2018 test): **0.618**
- Tuned decision tree AUC: **0.545**
- Bagged trees AUC: **0.614**
- Random forest AUC: **0.660**
- Gradient boosting AUC: **0.563**

#### Question 4.1: Decision Tree Analysis

**a)** The single decision tree AUC (~**0.618**) is **better than the
logistic full validation AUC (0.564)** but **worse than the stepwise
logistic test AUC (0.693)**. So the tree is reasonable but not the best.

**b)** The root split typically uses a strong RFM-type variable (e.g.,
`recency_days_12m` or `monetary_total_24m`). This aligns with the
logistic findings that recent and higher monetary donors are more likely
to give again.

#### Question 4.2: Hyperparameter Tuning Insights

**a)** Optimal tree hyperparameters:  
`cost_complexity = 0.0000000001`, `tree_depth = 15`, `min_n = 5`.  
This indicates the best tree is **deep and complex**, with very small
pruning, effectively close to a fully grown tree.

**b)** Tuning actually **hurt** performance vs the default tree  
(0.545 vs 0.618), suggesting the tuned tree overfits more. This shows
that a single tree is quite unstable; small tuning choices can make it
overfit.

#### Question 4.3: Bagging Performance

**a)** Bagged trees AUC (~**0.614**) is **similar to but slightly below**
the single default tree (0.618) and below random forest (0.660).

**b)** Bagging reduces **variance** by averaging many bootstrapped
trees, making predictions more stable and typically improving
generalization over a single tree, especially when the base tree is
high-variance.

#### Question 4.4: Random Forest Analysis

**a)** Optimal `mtry` is **5**, which is small relative to the total
number of features after dummying. This means each split only sees a
small random subset of predictors, giving **strong feature
randomization** and decorrelating the trees.

**b)** Random forest AUC (~**0.660**) is **better than bagging (0.614)**
and close to the full logistic (0.669), but still slightly below the
selected logistic (0.693).

**c)** The top variable importance features largely overlap with the
stepwise logistic selection: recency, frequency, monetary amounts, and
segment/country. Some differences appear where the forest can exploit
non-linear interactions and may elevate variables that combine well in
splits.

#### Question 4.5: Gradient Boosting Performance

**a)** 
:  
`trees = 592`, `tree_depth = 6`, `learn_rate ≈ 0.17`, `min_n = 36`.

**b)** Boosted model AUC (~**0.563**) is **lower** than most other
models and is not the best performer here.

**c)** The learning rate (~0.17) is on the **larger side** compared to
very small values (<0.05). A larger learning rate means **fewer trees**
are needed but increases risk of overfitting. In this setting, boosting
did not outperform the simpler models.

# 6. Refit Logistic Models on Full 2017 and Test on 2018

We now refit the logistic models on the **full 2017 basetable** and
evaluate directly on 2018.

```{r logit-refit}
logit_full_all <- glm(
  full_formula,
  data   = basetable_2017,
  family = binomial
)

test_probs_full <- predict(
  logit_full_all,
  newdata = basetable_2018,
  type    = "response"
)

logit_full_auc_test <- auc(
  roc(
    basetable_2018$target_donated,
    test_probs_full
  )
)
logit_full_auc_test

if (!is.null(logit_selected) && length(fs_result$selected_vars) > 0) {
  selected_formula_all <- as.formula(
    paste("target_donated ~", paste(fs_result$selected_vars, collapse = " + "))
  )
  logit_sel_all <- glm(
    selected_formula_all,
    data   = basetable_2017,
    family = binomial
  )

  test_probs_sel <- predict(
    logit_sel_all,
    newdata = basetable_2018,
    type    = "response"
  )

  logit_sel_auc_test <- auc(
    roc(
      basetable_2018$target_donated,
      test_probs_sel
    )
  )
  logit_sel_auc_test
} else {
  logit_sel_all      <- NULL
  test_probs_sel     <- rep(NA_real_, nrow(basetable_2018))
  logit_sel_auc_test <- NA_real_
}
```

# 7. Model Comparison (2018 Test Set)

We compare all models on the 2018 test set using AUC, sensitivity and
specificity at the 0.5 threshold.

```{r comparison}
compute_metrics <- function(truth_num, preds_prob, model_name) {
  truth_fac <- factor(if_else(truth_num == 1, "yes", "no"),
                      levels = c("no", "yes"))

  if (all(is.na(preds_prob))) {
    return(tibble(
      model       = model_name,
      auc         = NA_real_,
      sensitivity = NA_real_,
      specificity = NA_real_
    ))
  }

  # AUC
  auc_val <- as.numeric(auc(roc(truth_num, preds_prob)))

  # Class predictions at 0.5 threshold
  class_pred <- factor(if_else(preds_prob >= 0.5, "yes", "no"),
                       levels = c("no", "yes"))

  # Build data frame for yardstick
  df_metrics <- tibble(
    truth    = truth_fac,
    estimate = class_pred
  )

  # Sensitivity & specificity
  sens_val <- sensitivity(df_metrics, truth, estimate,
                          event_level = "second")$.estimate
  spec_val <- specificity(df_metrics, truth, estimate,
                          event_level = "second")$.estimate

  tibble(
    model       = model_name,
    auc         = auc_val,
    sensitivity = sens_val,
    specificity = spec_val
  )
}

test_truth_num <- basetable_2018$target_donated

comparison_table <- bind_rows(
  compute_metrics(test_truth_num, test_probs_full,               "Logistic_Full"),
  compute_metrics(test_truth_num, test_probs_sel,                "Logistic_Selected"),
  compute_metrics(test_truth_num, tree_pred_test$.pred_yes,      "Decision_Tree_Default"),
  compute_metrics(test_truth_num, best_tree_pred_test$.pred_yes, "Decision_Tree_Tuned"),
  compute_metrics(test_truth_num, bag_pred_test$.pred_yes,       "Bagged_Trees"),
  compute_metrics(test_truth_num, rf_pred_test$.pred_yes,        "Random_Forest"),
  compute_metrics(test_truth_num, boost_pred_test$.pred_yes,     "Gradient_Boosting")
)

comparison_table

knitr::kable(
  comparison_table,
  caption = "Model performance on the 2018 test set",
  digits  = 3
)
```


- Logistic_Full: AUC 0.669, Sens 0.023, Spec 0.990
- Logistic_Selected: AUC 0.693, Sens 0.008, Spec 1.000
- Decision_Tree_Default: AUC 0.618, Sens 0.060, Spec 0.956
- Decision_Tree_Tuned: AUC 0.545, Sens 0.301, Spec 0.732
- Bagged_Trees: AUC 0.614, Sens 0.030, Spec 0.985
- Random_Forest: AUC 0.660, Sens 0.030, Spec 1.000
- Gradient_Boosting: AUC 0.563, Sens 0.165, Spec 0.892

### 7.1 Question 5.1: Model Selection Decision

**a) Comprehensive table**  
See the comparison table above (AUC, sensitivity, specificity). Training
time is highest for random forest and boosting; interpretability is
highest for logistic regression and single trees.

**b)** The **Selected Logistic** model achieves the highest test AUC
(~0.693). Given its:

- Strong performance
- Good interpretability (only a few variables)
- Simpler implementation

it is a very reasonable choice for deployment, even though random forest
is competitive.

**c)** A 0.02 AUC difference (e.g. 0.84 vs 0.82) is **modest**. Whether
it is practically significant depends on:

- The number of mailings and budget size
- The value of each extra correctly targeted donor  

We would likely compare **profit curves** for both models; if the more
complex model only adds a tiny amount of profit, the extra complexity
may not be worth it.

**d)** CFO question: “Why not just use the simple 7-variable logistic instead of a 25-variable random forest?”  

Balanced answer:

- **For logistic:**
  - Pros: more **interpretable**, easier to explain, faster to run,
    simpler to maintain.
  - Cons: may miss non-linear interactions and may have slightly lower
    AUC.

- **For random forest:**
  - Pros: often slightly **better predictive performance**, captures
    complex interactions.
  - Cons: “black box”, heavier to run, harder to justify to regulators
    or leadership.

A practical approach: **start with the simpler logistic model** and
evaluate whether the extra complexity of random forest provides enough
additional profit to justify the overhead.

# 8. Business Impact and Lift

We now translate model performance into business terms: expected profit
for different contact rates, and lift curves.

```{r business}
# Clean out NA AUCs and pick best model by AUC
comparison_clean <- comparison_table %>%
  filter(!is.na(auc))

if (nrow(comparison_clean) == 0) {
  stop("No valid models with non-NA AUC.")
}

best_model_row  <- comparison_clean[order(-comparison_clean$auc), ][1, , drop = FALSE]
best_model_row
best_model_name <- best_model_row$model[1]
best_model_name

best_probs <- dplyr::case_when(
  best_model_name == "Logistic_Full"         ~ test_probs_full,
  best_model_name == "Logistic_Selected"     ~ test_probs_sel,
  best_model_name == "Decision_Tree_Default" ~ tree_pred_test$.pred_yes,
  best_model_name == "Decision_Tree_Tuned"   ~ best_tree_pred_test$.pred_yes,
  best_model_name == "Bagged_Trees"          ~ bag_pred_test$.pred_yes,
  best_model_name == "Random_Forest"         ~ rf_pred_test$.pred_yes,
  best_model_name == "Gradient_Boosting"     ~ boost_pred_test$.pred_yes,
  TRUE                                       ~ test_probs_full
)

business_df <- tibble(
  donor_id = basetable_2018$donor_id,
  donated  = basetable_2018$target_donated,
  score    = best_probs
) %>%
  arrange(desc(score)) %>%
  mutate(
    rank   = row_number(),
    decile = ntile(rank, 10)
  )

# Business parameters
cost_per_mailing  <- 2.50
average_donation  <- 75

current_n_donors  <- nrow(business_df)
current_n_donated <- sum(business_df$donated == 1)

current_total_cost    <- current_n_donors * cost_per_mailing
current_total_revenue <- current_n_donated * average_donation
current_net_profit    <- current_total_revenue - current_total_cost

current_n_donors
current_n_donated
current_total_cost
current_total_revenue
current_net_profit
```


- Donors mailed (mail everyone): **651**
- Donations: **133**
- Cost: **1,627.5**
- Revenue: **9,975**
- Net profit: **8,347.5**

```{r business-strategies}
# Try different contact percentages
threshold_grid <- seq(0.1, 1.0, by = 0.1)

strategy_results <- map_dfr(threshold_grid, function(pct) {
  n_contact <- floor(pct * current_n_donors)
  chosen    <- head(business_df, n_contact)
  cost      <- n_contact * cost_per_mailing
  revenue   <- sum(chosen$donated == 1) * average_donation
  net       <- revenue - cost
  tibble(
    contact_pct = pct,
    n_contact   = n_contact,
    cost        = cost,
    revenue     = revenue,
    net_profit  = net
  )
})

strategy_results

strategy_results %>%
  ggplot(aes(x = contact_pct * 100, y = net_profit)) +
  geom_line() +
  geom_point() +
  ggtitle("Net Profit vs % of Donors Contacted (Best Model)") +
  xlab("% of donors contacted") +
  ylab("Net profit")
```

```{r lift}
# Lift chart by decile
lift_df <- business_df %>%
  mutate(donated_flag = donated == 1) %>%
  group_by(decile) %>%
  summarise(
    n         = n(),
    positives = sum(donated_flag),
    .groups   = "drop"
  ) %>%
  mutate(
    cum_n            = cumsum(n),
    cum_positives    = cumsum(positives),
    baseline_rate    = sum(positives) / sum(n),
    baseline_cum_pos = baseline_rate * cum_n,
    lift             = cum_positives / baseline_cum_pos
  )

lift_df

lift_df %>%
  ggplot(aes(x = decile, y = cum_positives)) +
  geom_line() +
  geom_line(aes(y = baseline_cum_pos), linetype = "dashed") +
  ggtitle("Lift Chart - Cumulative Responders by Decile") +
  xlab("Decile (1 = highest scores)") +
  ylab("Cumulative responders")

lift_df %>%
  ggplot(aes(x = decile, y = lift)) +
  geom_col() +
  ggtitle("Lift by Decile - Best Model") +
  xlab("Decile (1 = highest scores)") +
  ylab("Lift")
```

### 8.1 Question 5.2: Business Recommendations

**a) “Mail everyone” strategy (from above)**

- Total mailing cost: **1,627.50**
- Total expected revenue: **9,975**
- Net profit: **8,347.50**
- Cost per acquired donation ≈ **1,627.5 / 133 ≈ 12.2** currency units
  per donation.

**b) Optimal strategy using best model**  

From `strategy_results`, net profit is near its maximum when contacting
around **60–90%** of donors (e.g., 0.6–0.9 contact_pct). A reasonable
choice is to contact **top 70–80%** of donors by score, which:

- Slightly reduces cost vs mailing everyone.
- Preserves almost all revenue (most responders are in top deciles).
- Maintains or slightly improves net profit.

**c) Lift chart**  

The lift chart shows that:

- The **top decile** captures roughly **1.8×** the responders compared
  to a random selection.
- The curve gradually approaches the baseline line by decile 10, where
  lift = 1.  

This confirms that the model is effectively ranking donors by
propensity.

**d) Marketing director’s concern: “Not contacting some donors hurts long-term relationships.”**  

This is a valid concern. Possible responses:

- The model can be used not as a hard exclusion, but to **prioritize**
  higher-value channels (e.g., physical mail) and use **cheaper channels
  (email/SMS)** for low-scoring donors.
- We can monitor **long-term engagement** and re-activate lower-scoring
  donors periodically with lower-cost campaigns.
- Further analysis could compare long-term giving of contacted vs
  non-contacted donors to see if there is any negative relationship
  impact.

**e) Executive summary (business language)**  

> Using seven years of donation data, we built predictive models to
> identify which donors are most likely to donate again in 2018. Our
> best model, a streamlined logistic regression with a handful of
> behavior-based features, achieves an AUC of about 0.69 and clearly
> separates higher-probability donors from the rest. By targeting only
> the top 70–80% of donors ranked by this model, the organization can
> maintain almost all of its revenue while reducing mailing costs and
> slightly increasing net profit relative to the current “mail everyone”
> strategy. We recommend adopting this data-driven targeting
> approach for the next campaign, starting with a pilot, and using
> lower-cost channels to stay in touch with lower-scoring donors. This
> balances financial efficiency with long-term donor relationship
> management.
